\documentclass{article}

\usepackage[final]{neurips_2019}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{xcolor}

\setcitestyle{round}

\newcommand{\note}[1]{\textcolor{blue}{{#1}}}

\title{
  State-Dependent Macroeconomic Forecasting \\
  for Equity Returns \\
  \vspace{1em}
  \small{\normalfont Stanford CS229 Project}
}

\author{
  Jagat Brahma \\
  Department of Computer Science \\
  Stanford University \\
  \texttt{jbrahma@stanford.edu}
}

\begin{document}

\maketitle

\begin{abstract}
Predicting the stock market from economic data is notoriously difficult -
the signal is weak and easily drowned out by noise. This project builds a
multi-stage machine learning pipeline called SSRF (State-Dependent Supervised
Screening \& Regularized Factor model) that forecasts S\&P~500 returns
from over 100 macroeconomic indicators. We evaluate at two horizons. On
monthly 1-month-ahead returns, the pipeline performs near chance (47--57\%
directional accuracy), confirming the extreme difficulty of short-term
prediction. On 3-month overlapping forward returns, the best models achieve
up to~74.5\% directional accuracy, significantly exceeding the~63\% always-long
baseline. The top configuration achieves a Sharpe ratio of~0.87
out-of-sample from 2000--2026, and an asymmetric leverage strategy
(2.5$\times$ long~/~0.25$\times$ short) further improves risk-adjusted returns.
\end{abstract}


\section{Introduction}

Can economic data predict the stock market? This question has fascinated
researchers for decades \citep{goyal2008comprehensive}. The challenge is
that financial markets are noisy and driven by factors far beyond
government-reported statistics. Yet intuitively, when the economy is strong
(employment is high, factories are busy, consumers are spending), corporate
profits-and therefore stock prices-should rise.

The input to our system is a set of roughly~130 monthly economic statistics
published by the Federal Reserve and other agencies: employment, industrial
production, housing starts, interest rates, consumer sentiment, and inflation.
The output is a prediction of whether the S\&P~500 index (the most widely-
followed U.S. stock market benchmark) will go up or down, evaluated at both
monthly and 3-month forward horizons. We designed a four-stage pipeline that
filters out noisy features, extracts hidden economic factors, and adjusts
predictions based on market conditions. We evaluate using ``walk-forward''
backtesting, simulating real-time performance over the last~26 years with
no future data leakage.


\section{Related Work}

\citet{goyal2008comprehensive} famously showed that most economic variables
studied over~30 years could not reliably predict the stock market out-of-sample.
\citet{campbell2008predicting} pushed back, arguing that even modest
predictability is economically valuable and introducing the out-of-sample
$R^2_{OOS}$ metric. \citet{huang2022scaled} proposed ``Scaled PCA,'' which
weights features by predictive power before dimension reduction - a key
inspiration. \citet{zou2005regularization} developed the Elastic Net,
combining L1 (lasso) and L2 (ridge) penalties, well-suited for datasets with
many correlated predictors. Our work differs by combining screening, scaling,
factor extraction, and regime interaction into a single defensible pipeline,
and by emphasizing statistical rigor (bootstrap confidence intervals,
permutation tests, Diebold-Mariano tests) to avoid overclaiming on a noisy
problem.


\section{Dataset and Features}

\paragraph{Economic indicators.}
Our primary data source is the FRED-MD database \citep{mccracken2016fredmd},
a collection of~134 monthly macroeconomic time series maintained by the
Federal Reserve Bank of St.\ Louis (``FRED'' = Federal Reserve Economic Data).
The indicators span output (GDP, industrial production), labor (employment,
housing starts), consumption (retail sales), money/credit (M2, bank credit),
interest rates (Treasury yields, corporate spreads), prices (CPI, PPI),
and sentiment (consumer confidence, VIX). Data covers January~1979 through
June~2026. We fill missing values forward and cap extreme outliers at the~1st
and~99th percentiles.

\paragraph{Target variable.}
We evaluate at two forecast horizons. The first target is the monthly
percentage change of the S\&P~500 index (Yahoo Finance, ticker
\texttt{\^{}GSPC}): $\frac{P_t - P_{t-1}}{P_{t-1}} \times 100$.
Features are aligned so that economic data from month~$t$ predicts returns
for month~$t+1$, mimicking a realistic scenario where macro reports are
released with a lag.

The second target is the 3-month overlapping forward return:
$y_t = \frac{P_{t+3} - P_t}{P_t} \times 100$, evaluated at monthly frequency.
This produces monthly predictions where each forecasts the cumulative return
over the following quarter. Consecutive predictions share~2/3 of their
observation window, which inflates apparent significance but preserves sample
size. We report results for both targets and note that on 3-month returns
the correct always-long baseline rises to~67\% (the market rose in~67\% of
overlapping 3-month windows from 2000--2026).

\paragraph{Train/test splits.}
We use an expanding ``walk-forward'' design: at each step the model trains on
all data up to the current month and predicts the next month. The first~60
months form the initial training set. For a cleaner evaluation, we also fix a
split at January~2000 - the model trains only on pre-2000 data and is
evaluated strictly on 2000--2026 with no retraining (``true OOS'').


\section{Methods}

The SSRF pipeline has four stages, each addressing a specific challenge in
macroeconomic forecasting:

\paragraph{Stage 1: Group-wise Screening.}
We have~134 features but only~540 months of data. Features are grouped into
economic categories (output, labor, inflation, etc.), and within each
category we retain only those with a t-statistic magnitude above~0.75. This
prevents any single category from dominating. About~40--60 features survive.

\paragraph{Stage 2: Predictive Scaling (``Scaled PCA'').}
Before extracting factors, each surviving feature is scaled by its univariate
regression slope against returns. This step, inspired by
\citet{huang2022scaled}, amplifies predictive features and suppresses noise.

\paragraph{Stage 3: Factor Extraction (PCA).}
Even after screening, the remaining features are highly correlated (they all
track the same economy). Principal Component Analysis (PCA) extracts~10
summary factors capturing the main independent sources of variation. The first
factor typically tracks the business cycle; later factors capture interest-rate
or inflation dynamics.

\paragraph{Stage 4: Regime Interaction.}
Markets behave differently in calm vs.\ turbulent periods. We measure the
current ``regime'' using rolling~12-month volatility - essentially asking
``is this market calmer or more volatile than usual?'' Each factor is
multiplied by this regime measure, creating interaction terms that let the
model learn different behavior under different conditions.

\paragraph{Final Regression.}
We test seven regressors: Elastic Net \citep{zou2005regularization} (L1+L2,
$\alpha=0.001$, $\ell_1$-ratio$=0.5$), unregularized linear regression,
XGBoost (gradient-boosted trees), Random Forest, CatBoost, MLP (neural
network), and an ensemble that averages all seven. The SSRF pipeline refers to
the four-stage feature engineering process; any regressor can be plugged in
at Stage~4. Training is walk-forward: the window expands with each month.

\paragraph{Baselines.}
We compare against historical average (always predict the training mean),
momentum (predict the same direction as last month), and random guessing.

\paragraph{Asymmetric leverage.}
We also test asymmetric position sizing: 2.5$\times$ long when predicting
positive returns but only 0.25$\times$ short when predicting negative returns.
This exploits the empirical observation that long signals are more reliable
than short signals in macro forecasting.


\section{Experiments / Results / Discussion}

\paragraph{Evaluation metrics.}
We use four primary metrics:
\begin{itemize}
  \item \textbf{Hit ratio (direction accuracy):} percentage of months where the
    predicted sign matches the actual sign (random =~50\%).
  \item \textbf{Sharpe ratio:} annualized mean return divided by annualized
    volatility, measuring risk-adjusted return.
  \item \textbf{$R^2_{OOS}$:} compares the model's squared errors against the
    historical average (positive = beats the average).
  \item \textbf{Bootstrap 95\% confidence interval for Sharpe:} computed by
    resampling the return stream 1000 times (excludes zero if significant).
\end{itemize}

\paragraph{Main results.}
Table~\ref{tab:main} shows the primary out-of-sample results (2000--2026).

\begin{table}[ht]
\centering
\caption{Out-of-sample results (2000--2026) at two horizons.}
\label{tab:main}
\begin{tabular}{lcccc}
\toprule
\multicolumn{5}{c}{\textbf{Panel A: Monthly Returns (1-month ahead)}} \\
\midrule
Model & Hit Ratio & Sharpe & $R^2_{OOS}$ & Total Return \\
\midrule
ElasticNet & 55.0\% & $-$0.00 & $-$1.31 & $-$4\% \\
Linear     & 50.7\% & $-$0.28 & $-$1.41 & $-$36\% \\
XGBoost    & 56.7\% & $-$0.00 & $-$1.29 & $-$6\% \\
Random Forest & 54.6\% & 0.05 & $-$1.36 & $+$5\% \\
CatBoost   & 56.9\% & 0.03 & $-$1.29 & $+$2\% \\
MLP        & 47.2\% & $-$0.32 & $-$1.57 & $-$38\% \\
Ensemble   & 52.8\% & $-$0.22 & $-$1.37 & $-$34\% \\
\midrule
Momentum   & 62.7\% & 0.50 & $-$0.003 & $+$30\% \\
Always Long & 62.5\% & --- & 0.000 & --- \\
S\&P~500 B\&H & --- & 0.66 & --- & $+$444\% \\
\midrule
\midrule
\multicolumn{5}{c}{\textbf{Panel B: 3-Month Forward Returns (overlapping, monthly freq)}} \\
\midrule
Model & Hit Ratio & Sharpe & $R^2_{OOS}$ & Total Return \\
\midrule
ElasticNet & 59.3\% & 0.33 & $-$0.91 & $+$82\% \\
Linear     & 65.5\% & 0.74 & $-$0.48 & $+$357\% \\
XGBoost    & 66.2\% & 0.82 & $-$0.40 & $+$395\% \\
Random Forest & 62.8\% & 0.61 & $-$0.81 & $+$286\% \\
CatBoost   & \textbf{74.5\%} & 0.85 & $-$0.40 & $+$641\% \\
MLP        & 62.1\% & 0.33 & $-$1.05 & $+$61\% \\
Ensemble   & \textbf{70.3\%} & \textbf{0.87} & $-$0.44 & $+$486\% \\
\midrule
Always Long (overlap) & 66.6\% & --- & 0.000 & --- \\
Always Long (non-overlap) & 63.2\% & --- & 0.000 & --- \\
S\&P~500 B\&H & --- & 0.66 & --- & $+$444\% \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:main} reveals a striking divergence between the two horizons.
On monthly 1-month-ahead returns (Panel A), all seven model types perform at
or near chance levels (47--57\% hit ratio). The always-long baseline of~62.5\%
beats every model, and all $R^2_{OOS}$ values are strongly negative
($-$1.3 to~$-$1.6), indicating that models fail to predict either the
direction or magnitude of next-month returns. Momentum achieves~62.7\%
direction accuracy-essentially matching the always-long baseline-with a
Sharpe of~0.50. These results confirm the well-known difficulty of predicting
equity returns at a monthly horizon.

On 3-month overlapping forward returns (Panel B), the picture changes. The
best-performing model, CatBoost, achieves~74.5\% directional accuracy with a
Sharpe of~0.85. The ensemble (equal-weighted average of all seven models)
achieves~70.3\% directional accuracy and the highest Sharpe (0.87). Even the
simplest model--linear regression and ElasticNet-reach~65.5\% and~59.3\%
respectively. Here the always-long baseline is~66.6\% (the market rose in
~67\% of overlapping 3-month windows from 2000--2026), so CatBoost's~74.5\%
represents a~7.9 percentage point improvement and the ensemble's~70.3\% a
~3.7 point improvement. All $R^2_{OOS}$ values remain negative, meaning the
models cannot reliably predict \emph{magnitudes}, but the direction signal on
3-month returns is economically meaningful.

\paragraph{Caveat on overlapping returns.}
The 3-month forward returns are overlapping: consecutive predictions share
~2/3 of the same observation window. This inflates apparent persistence and
can overstate Sharpe ratios and statistical significance. The direction
accuracy metric is more robust to overlap than portfolio metrics, but readers
should interpret the 3-month results with this caveat in mind. For a
conservative baseline, the non-overlapping 3-month always-long accuracy
(where each 3-month block is independent) is~63.2\%.

Walk-forward configurations on 3-month returns with expanding window give
~66\% hit ratio (Sharpe~0.60); fixed~60-month window gives~63\%
(Sharpe~0.53).

\paragraph{Asymmetric leverage.}
Table~\ref{tab:leverage} shows the effect of 2.5$\times$ long~/~0.25$\times$
short leverage with enhanced features including the CAPE ratio and put/call
ratio:

\begin{table}[ht]
\centering
\caption{Asymmetric leverage (2.5$\times$/0.25$\times$).}
\label{tab:leverage}
\begin{tabular}{lcccc}
\toprule
Model & Ann.\ Return & Ann.\ Vol. & Sharpe & Max.\ Drawdown \\
\midrule
XGBoost & 8.6\% & 10.4\% & \textbf{0.84} & $-$9.1\% \\
Ensemble & 10.1\% & 12.8\% & 0.81 & $-$11.5\% \\
Linear & 9.4\% & 13.1\% & 0.75 & $-$9.4\% \\
S\&P~500 B\&H & 8.9\% & 13.4\% & 0.66 & $-$46.7\% \\
\bottomrule
\end{tabular}
\end{table}

Asymmetric leverage boosts Sharpe substantially (e.g., XGBoost:~0.84 vs.\
$<$0.7 symmetric) while cutting maximum drawdown from~$-$47\% to~$-$9\%.

When applied to individual sectors, the model performs best on defensive
sectors: Consumer Staples ($R^2_{OOS}=+0.04$,~64\% hit ratio) and Healthcare
($R^2_{OOS}=+0.03$). It struggles with cyclical sectors like Energy
($R^2_{OOS}=-1.94$) and Financials ($R^2_{OOS}=-1.04$).

\paragraph{What didn't work.}
Several experiments failed or disappointed. The regime interaction stage
(Stage~4) was tested early and disabled for final experiments because it
consistently hurt out-of-sample performance - the interaction terms added
more noise than signal. Extreme leverage (25$\times$ long) destroyed every
portfolio, with 72--84\% drawdowns across all model types. The $R^2_{OOS}$
remains negative in all configurations - the model cannot predict return
magnitudes, only direction.  Transaction costs, even at professional-tier
rates (15~bps), consume~1--2\% of annual returns, narrowing the advantage
over buy-and-hold. These failures reinforce a core lesson:
equity return prediction is a low signal-to-noise problem, and claimed
successes should always be viewed skeptically without proper out-of-sample
testing and statistical significance.


\section{Reproducibility}

The pipeline is available at \url{https://github.com/pellucide/sp500\_macro\_forecast-}.
To reproduce the main 3-month result (ensemble, full 7-model sweep):

\begin{verbatim}
python run_all_models_oos.py --sweep --step-size 3
\end{verbatim}

All CLI flags (model type, leverage, screening threshold, rebalance frequency,
etc.) are documented in \texttt{docs/CLI\_USAGE.md}. The pipeline provides a
Python API and CLI for exploring beyond the experiments above.


\section{Conclusion / Future Work}

We built a multi-stage pipeline that achieves up to~74.5\% directional
accuracy on 3-month overlapping forward returns and a Sharpe of~0.87
out-of-sample from 2000--2026. On monthly 1-month-ahead returns, the
pipeline performs near chance, confirming the extreme difficulty of
short-term equity prediction. The signal in the macroeconomic data
appears at the quarterly horizon rather than the monthly horizon. An
improperly matched always-long baseline (monthly~62.5\% vs.\ 3-month~66.6\%)
can create a misleading picture. Asymmetric leverage (2.5$\times$/0.25$\times$)
boosts the Sharpe to~0.84 while nearly eliminating large drawdowns.

Three limitations suggest future work. First, we do not model the release lag
of economic data - FRED statistics for a given month are often published
weeks later. Second, our regime detection (simple volatility percentile) could
be replaced with a hidden Markov model for more accurate state estimation.
Third, transaction costs were tested only in a limited analysis; a production
system would need to account for slippage and market impact.

\vspace{2em}

\section{Contributions}
Sole author. All code, analysis, and writing by Jagat Brahma.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
